---
title: "Economic insecurity, lack of representation and radical right voting in the context of the 2021 German federal election"
author: "Candidate number: 1050713"
date: "`r format(Sys.time(), '%d %B %Y')`"
header-includes: 
  - \usepackage[T1]{fontenc}
  - \usepackage{fouriernc}
  - \usepackage{setspace}\onehalfspacing
  - \usepackage{amsfonts}
  - \usepackage{dcolumn}
  - \usepackage{pifont}
  - \usepackage{booktabs} 
  - \usepackage{placeins}
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}  
  - \fancyhf{}  
  - \rhead{Final AVCD assignment}  
  - \lhead{1050713}
  - \renewcommand{\headrulewidth}{0pt}
  - \cfoot{\thepage}   
mainfont: fouriernc
output:
  pdf_document:
    fig_caption: true
    keep_tex: true
    includes:
      in_header: preamble.tex
html_document: default
word_document: default
citation_package: natbib
bibliography: references.bib 
link-citations: yes
linkcolor: gray
urlcolor: magenta
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


```{r preliminaries, message=FALSE, warning=FALSE}
# load relevant packages 
library(tidyverse)
library(haven)
library(modelsummary)
library(survey)
library(here)
library(ggeffects)
library(margins)
library(marginaleffects)
library(kableExtra)
library(knitr)

# import data 
gles <- read_dta(paste0(here(), "/Data/ZA7701_v2-0-0_en.dta"))
```



```{r data-wrangling}
gles_mod <- gles %>% 
  select(1:100, grep("d38|d4|q18|d63|d18|q63|d17|d8|d7|wum6|q13|q14|q15|q16|q18|q23|q24|q25|q26||q27q35|q37|q46|q78|q79|q125|q143", names(gles))) %>%
  mutate(btw17_zweitstimme = ifelse(q34ba < 0, NA, q34ba),
         btw21_zweitstimme = ifelse(q114ba < 0, NA, q114ba),
         btw21_turnout = ifelse(q111 < 0 | q111 == 8, NA, q111),
         btw21_turnout1 = ifelse(btw21_turnout == 1, 1, 0),
         year_born = ifelse(grepl("-99|frueher", d2a), NA, d2a),
         ostwest2_dummy = ifelse(ostwest2 < 0, NA, ostwest2),
         ostwest_factor = factor(ostwest2_dummy, 
                                 levels = c(0, 1), 
                                 labels = c("ost", "west")),
         sex = ifelse(d1 < 0, NA, d1),
         sex1 = factor(sex, 
                       levels = c(1, 2), 
                       labels = c("male", "female")),
         year_born1 = as.numeric(as.character(year_born)),
         age = 2021 - as.numeric(as.character(year_born)),
         spd_21 = ifelse(btw21_zweitstimme == 4, 1, 0),
         union_21 = ifelse(btw21_zweitstimme == 1, 1, 0),
         gruene_21 = ifelse(btw21_zweitstimme == 6, 1, 0),
         fdp_21 = ifelse(btw21_zweitstimme == 5, 1, 0),
         afd_21 = ifelse(btw21_zweitstimme == 322, 1, 0),
         linke_21 = ifelse(btw21_zweitstimme == 7, 1, 0), 
         spd_to_switch = ifelse(btw21_zweitstimme == 4 & btw17_zweitstimme != 4, 1, 0),
         afd_away_switch = ifelse(btw17_zweitstimme == 322 & btw21_zweitstimme != 322, 1, 0),
         constituency_centric_rep = ifelse(q63a < 0, NA, q63a),
         party_centric_rep = ifelse(q63c < 0, NA, q63c),
         household_income = ifelse(d63 < 0, NA, d63),
         household_income_factor = as.factor(household_income),
         bachelor_dummy = ifelse(d8j1 < 0, NA, d8j1),
         school = ifelse(d7 < 0, NA, d7),
         abitur = ifelse(d7 == 5, 1, 0),
         abitur_factor = ifelse(abitur == 1, "abitur", "no_abitur"),
         urban_rural = ifelse(wum6 < 0, NA, wum6),
         urban_rural_factor = as.factor(urban_rural),
         subjective_class = ifelse(d38 < 0, NA, d38),
         left_right_self = ifelse(q37 < 0, NA, q37),
         left_right_self_factor = as.factor(left_right_self),
         left_right_cdu = ifelse(q35b < 0, NA, q35b),
         left_right_cdu_factor = as.factor(left_right_cdu),
         distance_cdu = (left_right_cdu-left_right_self)^2,
         left_right_csu = ifelse(q35c < 0, NA, q35c),
         left_right_csu_factor = as.factor(left_right_csu),
         distance_csu = (left_right_csu-left_right_self)^2,
         left_right_spd = ifelse(q35d < 0, NA, q35d),
         left_right_spd_factor = as.factor(left_right_spd),
         distance_spd = (left_right_spd-left_right_self)^2,
         left_right_afd = ifelse(q35h < 0, NA, q35h),
         left_right_afd_factor = as.factor(left_right_afd),
         distance_afd = (left_right_afd-left_right_self)^2,
         left_right_fdp = ifelse(q35e < 0, NA, q35e),
         left_right_fdp_factor = as.factor(left_right_fdp),
         distance_fdp = (left_right_fdp-left_right_self)^2,
         left_right_green = ifelse(q35f < 0, NA, q35f),
         left_right_green_factor = as.factor(left_right_green),
         distance_green = (left_right_green-left_right_self)^2,
         left_right_linke = ifelse(q35g < 0, NA, q35g),
         left_right_linke_factor = as.factor(left_right_linke),
         distance_linke = (left_right_linke-left_right_self)^2,
         scholz_love = ifelse(q18b < 0, NA, q18b),
         scholz_love_factor = as.factor(scholz_love),
         finzanz_abgehangt_subjektiv = ifelse(q46a < 0, NA, q46a),
         finzanz_abgehangt_subjektiv_factor = as.factor(finzanz_abgehangt_subjektiv),
         arbeit_abgehant_subjektiv = ifelse(q46b < 0, NA, q46b),
         arbeit_abgehant_subjektiv_factor = as.factor(arbeit_abgehant_subjektiv),
         cancel_culture_subjektiv = ifelse(q46d < 0, NA, q46d),
         cancel_culture_subjektiv_factor = as.factor(cancel_culture_subjektiv),
         infrastruktur_subjektiv = ifelse(q46c < 0, NA, q46c),
         infrastruktur_subjektiv_factor = as.factor(infrastruktur_subjektiv),
         unemployed_last10_yrs = ifelse(d17a < 0, NA, d17a),
         unemployed_last10yrs_months = ifelse(d17b < 0, NA, d17b),
         unemployed_last10yrs_weeks = ifelse(d17c < 0, NA, d17c),
         unemployed_dummy = ifelse(unemployed_last10_yrs != 0, 1, 0),
         unemployed_dummy_factor = as.factor(unemployed_dummy),
         trust_in_politicians = ifelse(q79d < 0, NA, q79d),
         trust_in_politicians_factor = as.factor(trust_in_politicians),
         trust_in_parliament = ifelse(q79b < 0, NA, q79b),
         trust_in_parliament_factor = as.factor(trust_in_parliament),
         trust_in_parties = ifelse(q79c < 0, NA, q79c),
         trust_in_parties_factor = as.factor(trust_in_parties),
         trust_in_public_broadcast = ifelse(q79i < 0, NA, q79i),
         trust_in_public_broadcast_factor = as.factor(trust_in_public_broadcast),
         trust_general = ifelse(q78 < 0, NA, q78),
         trust_general_factor = as.factor(trust_general),
         out_group_minorities_assim = ifelse(q125a < 0, NA, q125a),
         out_group_minorities_assim_factor = as.factor(out_group_minorities_assim),
         out_group_majority_will = ifelse(q125b < 0, NA, q125b),
         out_group_majority_will_factor = as.factor(out_group_majority_will),
         out_group_immig_econ_good = ifelse(q125c < 0, NA, q125c),
         out_group_immig_econ_good_factor = as.factor(out_group_immig_econ_good),
         out_group_immig_culture_threat = ifelse(q125d < 0, NA, q125d),
         out_group_immig_culture_threat_factor = as.factor(out_group_immig_culture_threat),
         out_group_immig_crime = ifelse(q125e < 0, NA, q125e),
         out_group_immig_crime_factor = as.factor(out_group_immig_crime),
         scale_pol_lasceht = ifelse(q18a < 0, NA, q18a),
         scale_pol_scholz = ifelse(q18b < 0, NA, q18b),
         scale_pol_baerbock = ifelse(q18c < 0, NA, q18c),
         econ_current_eval_general = ifelse(q23 < 0, NA, q23), 
         econ_current_eval_general_factor = as.factor(econ_current_eval_general),
         econ_current_personal = ifelse(q13 < 0, NA, q13), 
         econ_current_personal_factor = factor(econ_current_personal),
         econ_personal_gov_resp = ifelse(q15 < 0, NA, q15),
         gender_too_far = ifelse(q27g < 0, NA, q27g),
         gender_too_far_factor = factor(gender_too_far),
         job_loss_year_next2yrs = ifelse(d18 < 0, NA, d18),
         job_loss_year_next2yrs_factor = factor(job_loss_year_next2yrs),
         length_unemp_last10yrs_yrs = ifelse(d17a < 0, NA, d17a),
         length_unemp_last10yrs_mon = ifelse(d17b < 0, NA, d17b),
         unemp_at_least_one_year = ifelse(length_unemp_last10yrs_yrs >= 1, 1, 0),
         unemp_at_least_one_year_factor = factor(unemp_at_least_one_year),
         profession_loss_next2yrs = ifelse(d19 < 0, NA, d19),
         profession_loss_next2yrs_factor = factor(profession_loss_next2yrs),
         profession_current = ifelse(d11 < 0, NA, d11),
         type_of_emp_contract = ifelse(d13 < 0, NA, d13),
         difference_whos_gov = ifelse(q117 < 0, NA, q117),
         difference_whos_gov_factor = factor(difference_whos_gov),
         difference_who_votes = ifelse(q118 < 0, NA, q118),
         difference_who_votes_factor = factor(difference_who_votes))
```



# Introduction 

+ State research questions.

+ Structure of essay. 


# Motivation and relation to literature {#motivation}


2023 marks the tenth year of the Alternative für Deutschland’s (AfD) existence. Originally founded by a coterie of eurosceptic professors, the AfD ran on an anti-Euro platform in the 2013 federal election [@grimm_rise_2015], failing to clear the five-percent threshold for parliamentary representation only by a whisker.^[The [AfD won 4.7\%](https://www.bundeswahlleiterin.de/bundestagswahlen/2013/ergebnisse/bund-99.html) of the party votes (*Zweitstimmen*).] Starting in mid 2015, as the influx of refugees rose, the AfD increasingly morphed into a nationalistic, anti-immigration party [@arzheimer_alternative_2017; @arzheimer_how_2019; @cantoni_persistence_2020]. With the salience of immigration high during the 2017 federal election [@kellermann_immigration_2022], the [AfD won 12.6\%](https://www.bundeswahlleiterin.de/info/presse/mitteilungen/bundestagswahl-2017/34_17_endgueltiges_ergebnis.html) of party votes, making it the largest opposition party. The AfD’s entry into the *Bundestag* led to the debate about the determinants of radical right support – which had been ongoing among political scientists at least since the late 1990s [e.g. @kitschelt_radical_1995; @arzheimer_contextual_2009] – gaining greater attention in both other disciplines^[@guriev_political_2022 do an excellent job of summarising especially economists' contributions to the literature on the determinants of voting for populist parties.] and public discourse.


While much of the debate on the demand-side determinants of radical-right voting has focused on the relative importance of economic and cultural factors respectively,^[Those on the economic side of the debate mostly do not deny the importance of cultural factors, such as nativist attitudes - but attribute these attitudes to economic factors or shocks [e.g. @colantone_surge_2019; @colantone_backlash_2021]. The cultural camp [@art_myth_2022], by contrast, sees such attitudes, at least in part, as exogenous to economic shocks, instead, conceiving of them as the key background condition without which economic insecurity would not have given rise to the electoral success of populists [@margalit_economic_2019; @margalit_cultural_2022].] the theoretical framework I test below belongs to another, somewhat more nascent strand of the literature. This strand argues, inter alia, that the effect of economic grievances is likely moderated by (i) voters' trust in political institutions, notably parties and parliaments, and (ii) the degree to which they feel their preferences are represented in the party system [e.g. @dustmann_europes_2017; @eichengreen_populist_2018; @ivanov_economic_2023]. This set of theoretical hypotheses - though often not explicitly acknowledged by non-political-scientists - echo not only the work by @katz_cartel_2009 on the cartelisation of (Western) European party systems, but also more recent work stressing the importance of representational deficits in driving populist support [e.g. @manow_ent-demokratisierung_2020; @schafer_demokratische_2021; @zurn_how_2022; @silva_parties_2023]. More importantly, these hypotheses have been subjected to little empirical testing, with @ivanov_economic_2023 being a notable exception, as, for instance, @sonin_historical_2022 notes in his review of the work by @eichengreen_populist_2018. This essay is intended as a small step towards filling this gap in the literature. 

# Theory and case selection {#theory}

To the end, let me expatiate on the theoretical intuition outlined above. Like many economic-insecurity-centred theories of radical right support, the theoretical argument starts from the observation that economic shocks, such as automation [@boix_democratic_2019] or trade liberalisation [@autor_china_2013], create winners and losers. This assumption holds true not only for most Western European countries [@colantone_trade_2018; @milner_voting_2021], in general, but also for Germany, in particular [@dauth_rise_2014; @dauth_adjustment_2021]. Those losers, the argument goes on, who do not believe the current political system can address their grievances will then turn to populist radical right parties. 

By way of unpacking this argument, note, first, that losers may believe the political system to be incapable of redressing their economic grievances for two distinct reasons. First, they might deem all mainstream parties to be much of a muchness, meaning none of the supposedly different parties is seen by losers as representing their preferences. In that instance, losers are likely to think it does not matter which mainstream party they vote for, or which of those parties governs. Second, even if voters believe at least one mainstream party to endorse policies conducive to redressing their economic grievances, they might not trust any mainstream politician to follow through on these promises.^[This ‘trust’ channel is distinct from a ‘valence’ channel, where voters fundamentally trust politicians, but judge their levels of competence to differ [@green_politics_2017].] 


These two reasons also show why the theory predicts economic losers to turn to radical right parties, as opposed to left-wing parties, as a result of their grievances. Many other economic-insecurity-type arguments are unable to resolve this puzzle, which arises because the latter’s platforms tend to be significantly more pro-redistributive than the former’s platforms. Using the terminology by @linz_breakdown_1978, the reason is that losers, as @eichengreen_populist_2018 notes, regard left-wing parties as loyal to the political system and in cahoots with the other mainstream actors. So, their redistributive promises lack credibility. By contrast, radical right parties’ semi-loyalty or disloyalty to the system means that voting for them increase the likelihood that the political system, characterised by the mainstream cartel [@katz_cartel_2009], will be disrupted. 

In sum, this theoretical argument implies (at least)^[Naturally, the argument could also be taken to imply that (i) a perceived lack of representation and (ii) low trust increase the probability of voting for the radical right. Given the space constraints and my focus on how these two factors moderate the effects of economic security, I prescind from testing these two hypotheses.] three testable hypotheses.

+ H1: Economic insecurity, *ceteris paribus*, increases probability of voting for the radical right.
+ H2: The effect of economic insecurity on the probability of voting for the radical right is, *ceteris paribus*, stronger for those who believe their interests not be represented in the current party system. 
+ H3: The effect of economic insecurity is, *ceteris paribus*, stronger for those who are more distrustful of parties and/or parliament. 


Before setting out my empirical strategy, let me briefly discuss my two reasons for choosing the case of the 2021 German federal election. First, the 2021 election occurred in the wake of the “third wave” of the Covid-19 pandemic, which caused widespread financial insecurity among German households, with more than one-third of households reporting in early 2021 that they would be unable to cover unexpected expenses of €2.000 within one month and had already started tapping into their rainy-day funds [@cziriak_publication_nodate]. Secondly, the pandemic was, as in most other countries, a politically contentious period, causing deep divides over the desirability and efficacy of both non-pharmaceutical interventions, such as mask mandates, and vaccination. The *Querdenker* movement, in particular, used their protests to stress that their interests were not represented by any mainstream party. Both factors mean that economic insecurity and subjectively perceived representational deficits were likely salient in the minds of some voters, making this election a suitable case for testing the above hypotheses.



# Data, variables and operationalisation


To test the above hypotheses, I use the  *German Longitudinal Election Study's* (GLES) 2021 post-election survey^[The data for the 2021 *Nachwahlbefragung* are available [here](https://search.gesis.org/research_data/ZA7701).] - a high-quality survey of a cross-section of 3424 (quasi-)randomly selected individuals containing a rich set of questions about respondents' social and political attitudes as well as their socio-demographic characteristics. 




+ How do I operationalise economic insecurity? 
    + unemployed (actual experience) (`unemp_at_least_one_year`)
    + fear of job loss (`job_loss_year_next2yrs`)
    + fear of losing profession or having to change profession (`profession_loss_next2yrs`)
    + subjective evaluation of current economic situation 
    
+ How do I operationalise lack of representation? 
   + no difference who one is voting for 
   + no difference who governs



```{r data-summary-table}
data_summary_table <- tribble(~"Variable", ~"Operationalisation", ~"Survey item(s)",  
                              "Radical right voting", "dummy for AfD Zweitstimme", "q114ba = 322", 
                              "Economic insecurity", "dummy for unemployment in past ten years", "d17a-c", 
                              "", "fear of job loss", "d18",
                              "", "fear of losing or having to change profession", "d19", 
                              "Political system not responsive", "no difference which party governs", "q117",
                              "", "no difference which party one votes for", "q118", 
                              "Trust in political system", "trust in parliament", "q79b", 
                              "", "trust in parties", "q79c",
                              "", "trust in politicians", "q79d")
# table 
data_summary_table %>%
  kbl(caption = "Summary of variables and their operationalisation", booktabs = T) %>%
  kable_styling(latex_options = c("scale_down", "hold_position")) %>%
  pack_rows("dependent variable", 1, 1) %>%
  pack_rows("independent variables", 2, 9) %>%
  add_footnote(label = "Unemployment experience is, following Dauth et al (2021), defined as an individual having been unemployed for at least one year.", notation = "symbol") %>%
  add_footnote(label = "Source: Codebook for GLES Cross-Section 2021, Post-Election, ZA7701, Dataset Version v1.0.0.", notation = "symbol")
```



+ dependent variable -> binary, Why? 
    + Eichengreen focuses mainly on radical right, as Sonin notes
    + hence focus justified 

+ proxy for economic insecurity -> justification

   

# Methodology and Results 



+ Justify model specification.

+ DAG would be cool, but probably not possible.

+ Presentation of results + interpretation.

+ Caveats. 
  
  + not causal, correlational analysis
  


  
  
  
```{r simple-models-insecurity}
# simple models 
## job loss fear 
afd_job_loss_fear <- glm(afd_21 ~ job_loss_year_next2yrs + household_income + age + abitur_factor + sex1 + urban_rural_factor + ostwest_factor, 
                         family = binomial(link = "logit"), 
                         data = gles_mod)
## loss of profession 
afd_prof_loss_fear <- glm(afd_21 ~ profession_loss_next2yrs + household_income + age + abitur_factor + sex1 + urban_rural_factor + ostwest_factor, 
                          family = binomial(link = "logit"), 
                         data = gles_mod)
## unemployment experience 
afd_unemp_exp <- glm(afd_21 ~ unemp_at_least_one_year + household_income + age + abitur_factor + sex1 + urban_rural_factor + ostwest_factor, family = binomial(link = "logit"), 
                         data = gles_mod)
## econ current general situation 
afd_econ_current <- glm(afd_21 ~ econ_current_personal + household_income + age + abitur_factor + sex1 + urban_rural_factor + ostwest_factor, family = binomial(link = "logit"), 
                         data = gles_mod)

# modelplot 
modelplot(list(afd_job_loss_fear, afd_prof_loss_fear, 
               afd_unemp_exp, afd_econ_current)) +
  geom_vline(xintercept = 0)
```




```{r simple-models-lack-rep}
# simple models 
## difference gov
afd_diff_gov <- glm(afd_21 ~ difference_whos_gov + household_income + age + abitur_factor + sex1 + urban_rural_factor + ostwest_factor, family = binomial(link = "logit"), 
                         data = gles_mod)
## difference voting 
afd_prof_loss_fear <- glm(afd_21 ~ difference_who_votes + household_income + age + abitur_factor + sex1 + urban_rural_factor + ostwest_factor, family = binomial(link = "logit"), 
                         data = gles_mod)
# modelplot
modelplot(list(afd_diff_gov, afd_prof_loss_fear)) +
  geom_vline(xintercept = 0)
```


```{r interaction-models-lack-rep1}
# interaction models 
## difference gov
### job loss fear 
afd_diff_gov_int1 <- glm(afd_21 ~ difference_whos_gov*job_loss_year_next2yrs + household_income + age + abitur_factor + sex1 + urban_rural_factor + ostwest_factor, 
                         family = binomial(link = "logit"), 
                         data = gles_mod)
afd_diff_gov_int11 <- glm(afd_21 ~ difference_whos_gov_factor*job_loss_year_next2yrs_factor + household_income + age + abitur_factor + sex1 + urban_rural_factor + ostwest_factor, 
                         family = binomial(link = "logit"), 
                         data = gles_mod)

### profession loss fear
afd_diff_gov_int2 <- glm(afd_21 ~ difference_whos_gov*profession_loss_next2yrs + household_income + age + abitur_factor + sex1 + urban_rural_factor + ostwest_factor, 
                         family = binomial(link = "logit"), 
                         data = gles_mod)
### unemployment experience
afd_diff_gov_int3 <- glm(afd_21 ~ difference_whos_gov*unemp_at_least_one_year + household_income + age + abitur_factor + sex1 + urban_rural_factor + ostwest_factor, 
                         family = binomial(link = "logit"), 
                         data = gles_mod)
afd_diff_gov_int31 <- glm(afd_21 ~ difference_whos_gov_factor*unemp_at_least_one_year_factor + household_income + age + abitur_factor + sex1 + urban_rural_factor + ostwest_factor, 
                         family = binomial(link = "logit"), 
                         data = gles_mod)
plot_predictions(afd_diff_gov_int31, condition = list("difference_whos_gov_factor", "unemp_at_least_one_year_factor"))
### personal current situation
afd_diff_gov_int4 <- glm(afd_21 ~ difference_whos_gov*econ_current_personal + household_income + age + abitur_factor + sex1 + urban_rural_factor + ostwest_factor, 
                         family = binomial(link = "logit"), 
                         data = gles_mod)
# modelplot 
modelplot(list(afd_diff_gov_int1, afd_diff_gov_int2, 
               afd_diff_gov_int3, afd_diff_gov_int4)) +
  geom_vline(xintercept = 0)
```






```{r interaction-models-lack-rep2}
# interaction models 
## difference gov
### job loss fear 
afd_diff_gov_int5 <- glm(afd_21 ~ difference_who_votes*job_loss_year_next2yrs + household_income + age + abitur_factor + sex1 + urban_rural_factor + ostwest_factor, family = binomial(link = "logit"), 
                         data = gles_mod)
### profession loss fear
afd_diff_gov_int6 <- glm(afd_21 ~ difference_who_votes*profession_loss_next2yrs + household_income + age + abitur_factor + sex1 + urban_rural_factor + ostwest_factor, family = binomial(link = "logit"), 
                         data = gles_mod)
### unemployment experience
afd_diff_gov_int7 <- glm(afd_21 ~ difference_who_votes*unemp_at_least_one_year + household_income + age + abitur_factor + sex1 + urban_rural_factor + ostwest_factor, family = binomial(link = "logit"), 
                         data = gles_mod)
### personal current situation
afd_diff_gov_int8 <- glm(afd_21 ~ difference_who_votes*econ_current_personal + household_income + age + abitur_factor + sex1 + urban_rural_factor + ostwest_factor, family = binomial(link = "logit"), 
                         data = gles_mod)

# modelplot 
modelplot(list(afd_diff_gov_int5, afd_diff_gov_int6, 
               afd_diff_gov_int7, afd_diff_gov_int8)) +
  geom_vline(xintercept = 0)
```

  
  
# Conclusion 

+ I find only extremely weak support for Eichengreen's theory. 


# References 



\FloatBarrier


# Appendix 



## Descriptive statistics 





## Robustness checks 

